---
title: "1_diff_expr"
author: "Joshua Levy"
date: "7/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load libraries
```{r}
library(tidyverse)
library(brms)
library(emmeans)
library(bayestestR)
library(tidybayes)
library(comprehenr)
library(stringr)
```

# load data
```{r}
data<-readRDS("../../data/dsp_data.rds")
targets<-c("any_mets","Distant_Mets","ln_only","MLH1","mo(node_stage)")
normalizations<-c("igg","hk")
control.markers<-c(colnames(data$igg$expr %>% select(contains("IgG"))))
available.markers<-setdiff(names(data$igg$expr),control.markers)
```

# load options
```{r}
marker<-"CD8"
ratio.markers<-setdiff(c(c(available.markers),""),c(marker))
normalization<-"igg"
target<-"any_mets"
covar<-"no_covar"
covariates<-c("age","sex")
if (length(covariates)>0){covar<-"covar"}
cell.type<-"no_celltype"
if (length(control.markers)>0){cell.type<-"celltype"}
interactions<-c()
```

# run model
```{r}
if.else<-function(c_,x,y){if(c_){return(x)}; return(y)}
run.mod<-function(marker,norm="igg",target="distant_mets",mlh1_int=F,ratio.marker=""){
  control.text<-paste("+",control.markers,sep="",collapse="+")
  covariate.text<-paste("+",setdiff(covariates,c(target)),sep="",collapse="+")
  if (length(control.markers)==0){
    control.text<-""
  }
  if (length(covariates)==0){
    covariate.text<-""
  }
  dat<-cbind(data[[norm]]$pheno,data[[norm]]$expr)
  if (target%in%c("ln_only","Distant_Mets")){
    dat<-dat[dat[,target]==1 | dat[,"any_mets"]==0,]
  }
  ratio.text<-if.else(ratio.marker!="",paste("+offset(",ratio.marker,")",sep=""),"")
  mod<-brm(as.formula(paste(marker,"~",target,"*macro_annot",if.else(mlh1_int,"*","+"),"MLH1",control.text,covariate.text,ratio.text,"+(1|patient_id)",sep="")),data=dat,seed=42,cores=4,prior=prior(normal(0,5)),refresh=0,algorithm = "sampling",iter=2000,backend = "rstan", silent=T,future=T)
  mod$marker<-marker
  mod$mlh1_int<-mlh1_int
  mod$target<-target
  mod$ratio.marker<-ratio.marker
  return(mod)
}

return.res<-function(mod){
  x<-mod$marker
  target<-mod$target
  mlh1_int<-mod$mlh1_int
  ratio.marker<-mod$ratio.marker
  rg<-ref_grid(mod)
  if (grepl("node_stage",target)){target<-"node_stage"}
  em<-emmeans(rg,as.formula(paste(str_glue("~{target}|macro_annot{if.else(mlh1_int,'*MLH1','')}"))))
  cont<-contrast(em,method="pairwise")
  cont_posterior_macro <- gather_emmeans_draws(cont)[,c("contrast","macro_annot",if.else(mlh1_int,"MLH1",c()),".draw",".value")] %>% pivot_wider(names_from=c("contrast","macro_annot",if.else(mlh1_int,"MLH1",c())),values_from=".value")
  
  if (T){
    em<-emmeans(rg,as.formula(paste(str_glue("~macro_annot|{target}{if.else(mlh1_int,'*MLH1','')}"))))
  cont<-contrast(em,method="pairwise")
  cont_posterior_targ <- gather_emmeans_draws(cont)[,c("contrast",target,if.else(mlh1_int,"MLH1",c()),".draw",".value")] %>% pivot_wider(names_from=c("contrast",target,if.else(mlh1_int,"MLH1",c()),),values_from=".value")#,method="consec"
  
  res_<-rbind(as.data.frame(describe_posterior(cont_posterior_macro,ci=0.95))[-1,c("Parameter","Median","CI_low","CI_high","pd")] %>% mutate(pd=2*(1-pd)),
        as.data.frame(describe_posterior(cont_posterior_targ,ci=0.95))[-1,c("Parameter","Median","CI_low","CI_high","pd")] %>% mutate(pd=2*(1-pd)))
  }
  else {
    res_<-as.data.frame(describe_posterior(cont_posterior_macro,ci=0.95))[-1,c("Parameter","Median","CI_low","CI_high","pd")] %>% mutate(pd=2*(1-pd))
  }
  res_$gene<-x
  res_$ratio.gene<-ratio.marker
  res_$target<-target
  res_$mlh1_int<-mlh1_int
  return(list(res=res_,posterior_macro=cont_posterior_macro,posterior_target=cont_posterior_targ))
}

run.workflow<-function(marker,norm="igg",target="distant_mets",mlh1_int=F,ratio.marker=""){
  return(return.res(run.mod(marker,norm,target,mlh1_int,ratio.marker))$res)
}
```

# example run
```{r, message=FALSE, warning=FALSE}
library(future)
library(tictoc)
library(comprehenr)

plan(multisession,workers=12)
res_new<-do.call(rbind,future.apply::future_lapply(new.markers<-c("CD8","CD4","CD3","CD11c",""),function(x) run.workflow("CD66b",normalization,target,F,x),future.seed = 42L))
toc()

res_tmp<-run.mod("CD8",normalization,target,F,"CD66b")
return.res(res_tmp)$res
res_tmp$mlh1_int

tic("series")
plan(sequential)
res_new<-future.apply::future_lapply(c("CD8","CD4","CD3"),function(x) run.mod(x,normalization,target,T),future.seed = 42L)
toc()

tic("series_v2")
res_new<-to_list(for (marker in c("CD8","CD4","CD3")){run.mod(marker,normalization,target,T)})
toc()

names(res_new)<-c("CD8","CD4","CD3")
```

# let's check out results
```{r}
library(emmeans)
library(tidybayes)
library(bayestestR)
library(tidyverse)
library(bayesplot)

```

# return results
```{r}
results<-return.res(res_new$CD66b)
```

# results, differencecs in marker expression by target (e.g., mets vs no mets) given macro-architecture (intra, inter, away)
```{r}
results$res %>% filter(startsWith(Parameter,"0 - 1"))
```

```{r}
results$res %>% filter(startsWith(Parameter,"0 - 1"))
```

# results, differencecs in marker expression by macro-architecture (intra, inter, away), reported by target (e.g., mets vs no mets) 
```{r}
results$res %>% filter(!startsWith(Parameter,"1 - 0"))
```


# for command line for scaling
```{r}
args = commandArgs(trailingOnly=TRUE)
normalization<-normalizations[as.integer(args[1])]
target<-targets[as.integer(args[2])]
marker<-available.markers[as.integer(args[3])]
saveRDS(run.mod(marker,normalization,target),str_glue("./models/{normalization}--{target}--{marker}.rds"))
```

```{r}
library(future)
library(future.apply)
plan(multisession,workers=12)
args = commandArgs(trailingOnly=TRUE)
target<-targets[as.integer(args[1])]
marker<-available.markers[as.integer(args[2])]
mlh1_int<-as.integer(args[3])
res_new<-do.call(rbind,future_lapply(ratio.markers,function(x) run.workflow(marker,"igg",target,mlh1_int==1,x),future.seed = 42L))
saveRDS(res_new,str_glue("./results_v2/{normalization}--{target}--{marker}.rds"))
```

# running same analyses with lme4, not included in final study, opting for bayesian methods
```{r}
library(EnhancedVolcano)
ratio.tests<-list()
for (ratio.marker in setdiff(markers,c(control.markers))){
  emmeans.res<-list()
  # ratio.marker<-"CD45"
  for (marker in setdiff(markers,c(control.markers,ratio.marker))){
    control.text<-paste("+",control.markers,sep="",collapse="+")
    covariate.text<-paste("+",setdiff(covariates,c(target)),sep="",collapse="+")
    if (length(control.markers)==0){
      control.text<-""
    }
    if (length(covariates)==0){
      covariate.text<-""
    }
    dat<-cbind(data[[norm]]$pheno,data[[norm]]$expr)
    if (target%in%c("ln_only","Distant_Mets")){
      dat<-dat[dat[,target]==1 | dat[,"any_mets"]==0,]
    }
    mod<-lmer(as.formula(paste(marker,"~",target,"*macro_annot",control.text,covariate.text,"+offset(",ratio.marker,")+(1|patient_id)",sep="")),data=dat)
    cont<-emmeans(mod, pairwise~any_mets|macro_annot, data=dat)$contrasts
    summ.cont<-summary(cont)
    emmeans.res[[marker]]<-data.frame(marker=rep(marker,length(summ.cont$contrast)),
                        contrast=summ.cont$contrast,
                        macro=summ.cont$macro_annot,
                        estimate=summ.cont$estimate,
                        p=summ.cont$p.value
            )
  }
  
  emmeans.res.df<-as.data.frame(do.call(rbind,emmeans.res))
  ratio.tests[[ratio.marker]]<-emmeans.res.df
}

```


```{r}
markers
visualize.volcano<-function(ratio.marker="CD3",macro_="intra"){
  EnhancedVolcano(emmeans.mini<-ratio.tests[[ratio.marker]] %>% filter(contrast=="0 - 1"&macro==macro_) %>% mutate(log2FoldChange=estimate,pvalue=p),#-log10(
    lab = rownames(emmeans.mini),
    pCutoff = (5)*10^(-2),#*nrow(emmeans.mini)
    x = 'log2FoldChange',
    y = 'pvalue',
    pointSize = 4.0,
    labSize = 6.0,
    colAlpha = 1,
    legendPosition = 'right',
    legendLabSize = 12,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    FCcutoff = 0.4)
}

```

